A general and efficient representation of Ancestral Recombination Graphs

### 2022-08-10 paper.tex
New sections on efficiency and recombination

\section*{ARGs and SPRs}
\citet{wiuf1999ancestry,wiuf1999recombination} recast
the coalescent with recombination
as a stochastic process along the genome rather than backwards in time.
In this formulation we begin at one end of a chromosome by
sampling a tree under the standard coalescent, and then reason
about how this tree is modified by recombination events
as we move along the genome. Although this algorithm is
significantly more complicated to simulate than Hudson's
backwards in time approach, it has led to substantial
insights into the process of recombination itself
and to the sequentially Markov Coalescent (SMC)
approximations~\citep{mcvean2005approximating,marjoram2006fast}
that are central to many modern methods.

[Mention the
and ClonalFrame~\citep{didelot2007inference} approximation somewhere.]

Under the coalescent with recombination (or SMC) each local
tree along the genome is transformed into the next by
a ``subtree prune-and-regraft'' (SPR)
% Hein et al call it the ``subtree transfer'' operation but it's
% the same thing.
operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance
% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
% say this is wrong. But then Bordewich went ahead and said
% yes, actually it is NP hard.
and is NP-hard to
compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

[Discussion of SPRs in an EARG]

@@

The equivalence between an ARG and a set of local trees
separated by SPRs is often mentioned [e.g. x y z], but
it is important to note that while extracting local
trees from an ARG is straightforward, the converse
process of reconstructing precisely the same ARG from
these trees is more problematic. If the internal
nodes are consistently labelled and unary nodes are
included in the local trees (e.g., Fig~\ref{fig-simplification})
then it is easy to exactly reconstruct the ARG.
Otherwise, the problem is much more difficult and it
may not be possible to uniquely reconstruct the original ARG.
Without labels for the internal nodes, we are forced to
to reason about which subtrees remain the same and which
differ due to the SPR separating adjacent trees in order to
assign node identity. Without the internal unary nodes marking
recombinations, we must infer their existence. In
general, there will not be a unique solution to this problem,
and so there can be many possible ARGs compatible with
a given list of
% want to say "standard phylogenetic" trees here or something, but
% and they don't have to be binary (just arity > 1), but thought
% this would be familiar to readers?
% Jere: Could just say leaf-labelled "marginal trees", "ancestral trees",
% "local trees", or something like that.
leaf-labelled binary trees.

[ Cite \citep{deng2021distribution} somewhere here.]

% This text might go somewhere
% We may have nodes corresponding to genomes that are ancestral to
% the sample along their entire span
% (e.g.\ node $H$ in Fig.~\ref{fig-ancestry-resolution}),
% genomes that carry no genetic material ancestral to the sample at all
% (e.g.\ node $J$ in Fig.~\ref{fig-ancestry-resolution}),
% and every possibility along the continuum between these extremes.
% Clearly, we can only ever hope to infer the fraction of an ancestor's
% genome that is ancestral to the sample.

@@

@@

\begin{itemize}
\item [Recap quickly distinction bewtween aARG and gARG. Then]
    This, coupled with
    the explicit association of genome coordinates with edges, allows
    us to represent even the bewilderingly complex patterns of relatedness
    present in modern datasets.

>>>>

\section*{Relationship between ARGs and trees}

We have seen that the local tree for a given position on
the genome can be extracted from an ARG by starting at the
leaves and following the appropriate paths through the graph.
However, this tells us little about the relationship \emph{between}
adjacent trees, and what properties the overall emsemble of
local trees might have.
A recombination between two adjacent trees causes them to
differ by a ``subtree prune-and-regraft'' (SPR)
% Hein et al call it the ``subtree transfer'' operation but it's
% the same thing.
operation~\citep{hein1990reconstructing,song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance
% Hein 1996 say that SPR distance is NP-hard, but Allen 2001
% say this is wrong. But then Bordewich went ahead and said
% yes, actually it is NP hard.
and is NP-hard to
compute~\citep{hein1996complexity,allen2001subtree,bordewich2005computational}.

@@

The equivalence between an ARG and a set of local trees
separated by SPRs is often mentioned [e.g. x y z], but
it is important to note that while extracting local
trees from an ARG is straightforward, the converse
process of reconstructing precisely the same ARG from
these trees is more problematic. If the internal
nodes are consistently labelled and unary nodes are
included in the local trees (e.g., Fig~\ref{fig-simplification})
then it is easy to exactly reconstruct the ARG.
Otherwise, the problem is much more difficult and it
may not be possible to uniquely reconstruct the original ARG.
Without labels for the internal nodes, we are forced to
to reason about which subtrees remain the same and which
differ due to the SPR separating adjacent trees in order to
assign node identity. Without the internal unary nodes marking
recombinations, we must infer their existence. In
general, there will not be a unique solution to this problem,
and so there can be many possible ARGs compatible with
a given list of leaf-labelled binary trees.
This is the approach taken by the Espalier inference
method~\citep{rasmussen2022espalier}, which finds a
set of SPRs that reconcile independently inferred
trees from different regions of the genome
using.

\section*{Precision of recombination information}

% Not sure where this goes, but there are some critical points
% that we want to explain and not just throw out there in the
% discussion.
A critical distinction between the eARG and gARG encodings
is the level of precision about the time and location of
recombination events that is required. In an eARG, we are
required to be absolutely precise about every recombination
[explain and contrast with gARG, referring to examples]

For example, \citep{deng2021distribution} state that
``[the ARG] provides
all the information about the genealogical history of a sample,
including the locations of recombination events.''

% JK: This is very rough, just jotting down the basic ideas,
% will need substantial revision.
It is not clear that such complete information will always
(or indeed ever) be available from inferences. We may be
able to precisely identify the location and timing
of recent recombinations (particularly if detailed pedigree
information is available) but beyond this, the information
required simply isn't present. When sampling from a
process such as the Sequentially Markov
Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
like ARGweaver~\citep{rasmussen2014genome}, it is true that
we do obtain a fully resolved ARG with precise
information about recombination, but it is reasonable
to question how much information from the data is used
to support older recombinations, in particular.

At the very least, it is not reasonable for the \emph{data
structure} to require complete precision in order to
represent some less precise. This is a fundamental
advantage of the gARG encoding over the classical eARG
method: we are not forced to posit a precise cause (recombination
event) for every effect (difference between adjacent trees).
These differences in precision of recombination
information are illustrated in the ARGs estimated by different
methods, as we see in the next section.

@@

\section*{Efficiency of computation on ARGs}
The succinct tree sequence data structure (usually known as a ``tree sequence''
for brevity, but see later comment on confusion around this point)
is a concrete encoding of a gARG, as discussed in this article.
It was originally developed as part of the \texttt{msprime}
simulator~\cite{kelleher2016efficient} and subsequently ...

[The tree sequence encoding is very efficient for lots of things. Examples]

In retrospect, the choice of terminology was unfortunate, and has led to
significant confusion. [More]

@@

\begin{itemize}

\item Recap quickly distinction bewtween aARG and gARG.Then:
    This, coupled with
    the explicit association of genome coordinates with edges, allows
    us to represent even the bewilderingly complex patterns of relatedness
    present in modern datasets.
